---
title: "611projectreport"
date: "2024-12-07"
output: pdf_document
---


```{r}
library(tidyverse)
library(ggplot2)
library(pROC)
library(gbm)
```

We first load and clean up our data. Our main goals are to get rid of missaligned data and make sure the Latitude and Longitudes are properly formatted.
```{r}
nyc <- read.csv("source_data/nyc_languages.csv", sep=";")

#following lines go onto a new line which missaligns their columns
#219-222 columbian spanish
#514-517 jewish neo-armaric
#568-571 kham tibetan
#606-607 kyiron
#1275-6 jewish neo-armaric description too long and on wrong line

#remove missaligned rows/columns

nyc_cln <- nyc[-c(219:222, 514:517, 568:571, 606:607, 1275:1276),]


nyc_map <- nyc_cln %>% select(Language, World.Region, Language.Family, Latitude, Longitude) %>% filter(complete.cases(.)) %>% filter(Language != "", World.Region != "", World.Region != "367357", Latitude != "", Longitude != "", Language.Family != "")

#latitudes and longitudes have dots that cannot be read a numeric

# Replace extra dots with nothing (removes them)
nyc_map$Latitude <- gsub("\\.", "", nyc_map$Latitude)
nyc_map$Longitude <- gsub("\\.", "", nyc_map$Longitude)

#then reinsert the decimal
nyc_map$Latitude <- sapply(nyc_map$Latitude, function(x) paste0(substr(x, 1, 2), ".", substr(x, 3, nchar(x))))
nyc_map$Longitude <- sapply(nyc_map$Longitude, function(x) paste0(substr(x, 1, 3), ".", substr(x, 4, nchar(x))))


#then make data numeric
nyc_map$Latitude <- as.numeric(nyc_map$Latitude)
nyc_map$Longitude <- as.numeric(nyc_map$Longitude)
```


Next, we want to be able to visualize our data. We are mainly interested in how Latitude and Longitude affects the primary World Region of the observed language spoken. We plot it below.
```{r}
region.plt <- ggplot(nyc_map, aes(x = Latitude, y = Longitude, color = World.Region)) + geom_point()

region.plt
```

As we can see, most of the data is concentrated in the center (central Manhattan), and its hard to see any clear pattern.

Another way to interpret the data is to sort the World Region of the language from most to least popular.

```{r}
#sort the by frequency
freq_nyc <- nyc_map %>% group_by(World.Region) %>% tally() %>% arrange(desc(n)) %>% mutate(World.Region = factor(World.Region, World.Region)) %>% mutate(rank=1:nrow(.))

#plot
freq_plt <- ggplot(freq_nyc, aes(rank, n)) + geom_segment(aes(x=rank, xend=rank, y=0, yend=n)) + theme_minimal() + geom_text(aes(label = World.Region, y = n+20), angle = -90, size = 2)

freq_plt
```
While there are exceptions such as Western Africa and Caribbean, the most popular language World Regions seem to be some part of Asia or Europe. Thus we further investigate these two regions further.

We can sort the data by creating a new variable to classify if the language is from Asia or Europe and eliminate the observation if it is from neither.
```{r}
s <- nyc_map %>%
  filter(World.Region %in% c(
    "Southern Europe", "Eastern Europe", "Western Europe", "Northern Europe", "Southern Asia",
    "Southeastern Asia", "Eastern Asia", "Western Asia", "Central Asia")) %>%
  mutate(class = case_when(
    World.Region %in% c("Southern Europe", "Eastern Europe", "Western Europe", "Northern Europe") ~ "Europe",
    World.Region %in% c("Southern Asia", "Southeastern Asia", "Eastern Asia", "Western Asia", "Central Asia") ~ "Asia"
  ))
```


Next we run a PCA to see if we can visualize observations from Asia or Europe
```{r}
# Select only the relevant columns for PCA
pca_prep <- s %>% select(Latitude, Longitude)

# Apply PCA on the selected columns
pca <- prcomp(pca_prep, center = TRUE, scale. = TRUE)  # It's a good idea to center and scale data

# Convert the PCA result into a data frame
pca.data <- as.data.frame(pca$x)

# Add the 'World.Region' label to the PCA result for plotting
pca.data$label <- s$class

# Plot the first two principal components with coloring by 'World.Region'
pca <- ggplot(pca.data, aes(x = PC1, y = PC2, color = as.factor(label))) + geom_point() + labs(color = "World Region", )

pca
```
Although the center is still quite clustered and hard to distinguish, as we move outwards the trend becomes more visually clear. 

To test if the location of the observation can correctly classify whether the language is from Asia or Europe, we split the data into a 75/25 train test dataset and run a random forest. The resulting ROC is shown below.
```{r}
set.seed(123)

#have a binary variable for class
s <- s %>% mutate(class_bern = case_when(class == "Europe"~0, class == "Asia"~1))

train_ii <- runif(nrow(s)) < .75
train <- s %>% filter(train_ii)
test <- s %>% filter(!train_ii)
model <- gbm(class_bern ~ Latitude + Longitude, data=train, distribution = "bernoulli")

predict_part <- predict(model, newdata = test, type = "response")
roc_curve <- roc(test$class_bern, predict_part)

plot(roc_curve, main = paste("ROC: AUC =", round(auc(roc_curve), 4)))
```



We can rerun this model several times to see what kind of results from the ROC that we get.
```{r}
set.seed(123)
rocs <- c()

for (i in 1:100) {
  train_ii <- runif(nrow(s)) < .75
  train <- s %>% filter(train_ii)
  test <- s %>% filter(!train_ii)
  model <- gbm(class_bern ~ Latitude + Longitude, data=train, distribution = "bernoulli")
  predict_part <- predict(model, newdata = test, type = "response")
  roc_curve <- roc(test$class_bern, predict_part)
  rocs[i] <- round(auc(roc_curve), 4)
}

hist(rocs, main = "AUC Distribution")
summary(rocs)
```
Seeing as our results range from .653 to .822, it seems that our AUC is highly dependent on how our training and test data are split up. It would make sense that the outer points are much easier to predict, so if the test data has more points on the outskirts of the data, if would predict correctly more often. Similarly, if our test data is mostly from the center cluster, than it will predict much more poorly.










